---
title: "Bootcamp_R_Final"
output: html_document
date: "2023-09-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()
```

## Import data in the data file
```{r cars}
school = read.csv("data/nys_schools.csv");
counties = read.csv('data/nys_acs.csv')
```

## Data Exploration

You can also embed plots, for example:

```{r another chunk name}
summary(counties)  # Summarize numeric columns
str(counties)     # Display the structure of the data
head(counties)    # Show the first few rows of the data
```



### We did not find missing value in table Counties
```{r another chunk name}
# Check for missing values in the "column_name" column
missing_values <- is.na(counties)

# Count the number of missing values in each column
missing_counts <- colSums(missing_values)

print(missing_counts)

```



## we found some outliers in Total_enroll, per_Free_lunch, Per_reduced_lunch, per_lep
## mean_ela_score, and mean_math_score
```{r another chunk name}
summary(school)  # Summarize numeric columns
str(school)     # Display the structure of the data
head(school)    # Show the first few rows of the data
```
### We did not find missing value in table School
```{r another chunk name}
# Check for missing values in the "column_name" column
missing_values <- is.na(school)

# Count the number of missing values in each column
missing_counts <- colSums(missing_values)

print(missing_counts)

```


### we found some outliers in Total_enroll, per_Free_lunch, Per_reduced_lunch, per_lep
## mean_ela_score, and mean_math_score,  so we remove the outlier and save it into another table
```{r another chunk name}
# Remove outliers from multiple columns
cleaned_school <- school %>%
  filter(
    total_enroll >= 0,
    per_free_lunch >=0 & per_free_lunch <= 1,
    per_reduced_lunch >=0 & per_reduced_lunch <= 1,
    per_lep >=0 & per_lep <= 1,
    mean_ela_score >=0,
    mean_math_score >= 0
  )

```




# We will split into 3 groups using the quantile value of the column county_per_poverty

```{r another chunk name}
# Calculate quartiles (25th, 50th, and 75th percentiles) for "column_name"
quartiles <- quantile(counties$county_per_poverty, probs = c(0.33, 0.66))

# Print the quartile values

q33 = quartiles[1] #33th Percentile
q66 = quartiles[2] #66th Percentile

```

```{r another chunk name}
# Calculate quartiles (25th, 50th, and 75th percentiles) for "column_name"
counties <- counties %>%
  mutate(prop_group = case_when(
    county_per_poverty <= q33 ~ "low",
    county_per_poverty <= q66 & county_per_poverty > q33 ~ "Medium",
    county_per_poverty > q66 ~ "High"
  ))
```



# We will split into 3 groups using the quantile value of the column county_per_poverty




```{r another chunk name}
### 3.3 Standardized test scores

# Make calculation manually
scores_std <- cleaned_school %>%
  select(year, contains("score")) %>%
  group_by(year) %>%
  summarize(ela_mean = mean(mean_ela_score, na.rm=TRUE),
            math_mean = mean(mean_math_score, na.rm=TRUE),
            ela_sd = sd(mean_ela_score, na.rm=TRUE),
            math_sd = sd(mean_math_score, na.rm=TRUE))

# Create z-score columns
schools_all = inner_join(cleaned_school, scores_std, by="year")
schools_all = mutate(schools_all,
                     ela_z_score = (mean_ela_score-ela_mean)/ela_sd,
                     math_z_score = (mean_math_score-math_mean)/math_sd)

head(schools_all)
```






```{r another chunk name}
# Perform an inner join between df1 and df2 on the common_column
head(cleaned_school)
```

### we will do a inner join between the table 

```{r another chunk name}
# Perform an inner join between df1 and df2 on the common_column
merge_table <- inner_join(counties, schools_all, by = c("county_name", "year"))
head(merge_table)
```


```{r another chunk name}

# Group by the "Group" column and calculate summary statistics
summary_data <- merge_table %>%
  group_by(county_name) %>%
  summarize(
    Total_enrollment = sum(total_enroll),
    Mean_per_free_lunch = mean(per_free_lunch),
    Mean_per_reduced_lunch = mean(per_reduced_lunch),
    Mean_per_poverty = mean(county_per_poverty),
    Count = n()
  )

# Print the summary data
print(summary_data)
```

```{r another chunk name}
sorted_data <- summary_data %>%
  arrange(desc(Mean_per_poverty))  # Use 'desc' for descending order

# Select the top 5 rows
top_5 <- sorted_data %>%
  slice_head(n = 5)

top_5

```

```{r another chunk name}
r_sorted_data <- summary_data %>%
  arrange(Mean_per_poverty)  # Use 'desc' for descending order

# Select the top 5 rows
bot_5 <- r_sorted_data %>%
  slice_head(n = 5)

bot_5

```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
